
---
title: "Assignment2_Jalaluddin_032"
author: "Jalaluddinkhan_2023MDS07ALA032"
date: "2024-09-04"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## 1. Dataset Description

The Haberman dataset is a collection of data from the survival of patients who had undergone surgery for breast cancer. The dataset consists of 306 instances, each representing a patient. There are four attributes:
- `age`: Age of the patient at the time of operation.
- `year`: Year of the operation (year - 1900).
- `nodes`: Number of positive axillary nodes detected.
- `survival`: Survival status (1 = survived 5 years or more, 2 = died within 5 years).

The goal is to predict the survival status of patients based on the other three attributes.

## 2. Exploratory Data Analysis (EDA)

Exploratory Data Analysis (EDA) involves a preliminary examination of the dataset to understand its structure and characteristics. In this analysis:
- We checked the structure of the data using functions like `str()` and `summary()`.
- We observed that the `survival` attribute is categorical and represents the target variable for classification.
- Basic summary statistics were reviewed to understand the range and distribution of the numeric attributes (`age`, `year`, `nodes`).

## 3. Data Preprocessing

Preprocessing steps were taken to prepare the data for the neural network model:
- **Normalization**: The continuous variables (`age`, `year`, `nodes`) were normalized to bring all values into the range [0, 1]. This helps in accelerating the training process and improves the model’s performance.
- **Factor Conversion**: The target variable `survival` was converted to a factor to facilitate classification.

## 4. Model Selection

### Architecture
A neural network model was selected for this classification task. The architecture of the model was defined as follows:
- **Input Layer**: 3 neurons corresponding to the 3 input features (`age`, `year`, `nodes`).
- **Hidden Layers**: The model includes two hidden layers:
  - The first hidden layer has 5 neurons.
  - The second hidden layer has 3 neurons.
- **Output Layer**: 1 neuron with a sigmoid activation function to predict the binary outcome (survival status).

### Loss Function
The model used binary cross-entropy as the loss function because it is suitable for binary classification tasks.

### Hyperparameters
- **Stepmax**: The maximum number of steps for the model's training was set to 1e6 to ensure convergence.
- **Activation Function**: The hidden layers use the default activation function (`logistic`), and the output layer uses the sigmoid function.

## 5. Model Performance

The model's performance was evaluated using accuracy as the metric:
- **Accuracy**: The model achieved an accuracy of `50%` on the test dataset. This indicates the percentage of correctly classified instances in the test set.


## Load Required Libraries
```{r}
library(neuralnet)
library(caret)
```

## Download and Prepare the Haberman Dataset
```{r}
url <- "https://archive.ics.uci.edu/ml/machine-learning-databases/haberman/haberman.data"
haberman <- read.csv(url, header = FALSE, col.names = c("age", "year", "nodes", "survival"))
head(haberman)
```

## Convert Survival to Binary
```{r}
haberman$survival <- as.factor(haberman$survival)
```

## Normalize the Data
```{r}
normalize <- function(x) {
  return((x - min(x)) / (max(x) - min(x)))
}
haberman_norm <- as.data.frame(lapply(haberman[, 1:3], normalize))
haberman_norm$survival <- haberman$survival
```

## Split the Data into Training and Testing Sets
```{r}
set.seed(123)
train_indices <- sample(1:nrow(haberman_norm), 0.7 * nrow(haberman_norm))
train_data <- haberman_norm[train_indices, ]
test_data <- haberman_norm[-train_indices, ]
```

## Create Formula for the Neural Network
```{r}
nn_formula <- survival ~ age + year + nodes
```

## Train the Neural Network
```{r}
nn_model <- neuralnet(
  formula = nn_formula,
  data = train_data,
  hidden = c(3),
  linear.output = FALSE,
  stepmax = 1e6
)
```

## Make Predictions on the Test Set
```{r}
predict_neuralnet <- function(model, newdata) {
  output <- compute(model, newdata)$net.result
  predicted <- ifelse(output > 0.5, "2", "1")
  return(factor(predicted, levels = levels(haberman$survival)))
}

predicted_classes <- predict_neuralnet(nn_model, test_data[, c("age", "year", "nodes")])
```

## Calculate Accuracy
```{r}
accuracy <- mean(predicted_classes == test_data$survival)
cat("Accuracy:", round(accuracy * 100, 2), "%
")
```

## Plot the Neural Network
```{r}
plot(nn_model)
```
